---
title: "Week 7 · Computational Musicology and (Audio) Visualisation"
author: "John Ashley Burgoyne"
date: "15 February 2023"
output: 
  html_document:
    toc: true
    theme:
      heading_font:
        google: 
          family: Rajdhani
          wght: 700
      base_font:
        google: Fira Sans
      code_font:
        google: Fira Mono
      bg: "#FFFFFF"
      fg: "#212529" 
      primary: "#2b2bee"
      secondary: "#39d7b8"
      success: "#39d7b8"
      danger: "#fa5577"
      warning: "#ffb14c"
      info: "#0cc7f1"
---

```{r, echo = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

You can download the raw source code for these lecture notes [here](compmus2024-w07.Rmd).

## Course Meeting Plan

### Wednesday · 45 February · Lecture

- Course intro and demo portfolio (10 min)
- Lecture: What is computational musicology? (15 min)
- Breakout 1: Spotify features (15 min)
- Plenary Discussion: Breakout findings (10 min)
- Lecture: What is a spectrogram? (15 min)
- Demo: Sonic Visualiser (10 min)
- Lecture: What is good analytical design? (15 min)

If you want some distraction *after* the lecture, have a look at the [Sources of Good Data Visualisations](https://canvas.uva.nl/courses/37009/pages/sources-of-sometimes-good-data-visualisations?module_item_id=1523431) on Canvas.

### Wednesday · 15 February · Lab

If you have time before this lab, [try to connect your portfolio to Spotify](https://canvas.uva.nl/courses/37009/pages/connecting-your-portfolio-to-spotify?module_item_id=1523448). If you get stuck, your TA will help you during the lab session.

- Introductions (5 min)
- Review: RStudio, R Markdown, and Spotify (15 min)
- Breakout 2: 'Let It Go' (20 min)
- Plenary Discussion: Breakout findings (5 min)
- Demo: ggplot (15 min)
- Breakout 3: Digster decade comparison (20 min)
- Plenary Discussion: Breakout findings (10 min)

## Breakout 1: Spotify Features

Visit the [Spotify feature documentation](https://developer.spotify.com/documentation/web-api/reference/#/operations/get-audio-features). It is quite short -- and Spotify provides no other explanation of what their features mean.

For each feature, discuss with your group which of the four 'musical moments' you think Spotify uses to compute it:

  1. Pitch
  2. Volume
  3. Timbre
  4. Duration/Time
  
Some of them are obvious, and some of them are very much not!

Be prepared to discuss your conclusions with the class at the end of the breakout.

## Breakout 2: 'Let It Go'

If you interacted with any young children in your life a few years ago, you also surely interacted with the Disney hit 'Let It Go', from *Frozen*. More recently, 'We Don't Talk about Bruno' from *Encanto* is fighting to take its place. What about these songs made them so irresistibly popular? You and your group will try to use the Spotify API to see if there is anything unusual.

### Introduction

Every analysis you do for this class must start with the following two lines, which load the `tidyverse` and `spotifyr` packages.

```{r}
library(tidyverse)
library(spotifyr)
```

The `spotifyr` library allows you to load information by track, album, artist, or playlist. I very strongly recommend using playlists. 

All of the functions rely on Spotify URIs. You can copy them from spotify, but you need to delete the first parts (e.g., `spotify:track:`, `spotify:album:`, `spotify:artist:`, or `spotify:playlist:`) and keep only the alphanumeric nonsense at the end. Note also that `get_playlist_audio_features()` requires an extra blank string at the beginning, for historical reasons.

```{r}
juditha <- get_track_audio_features(c("2M5b9YLAgFroqWzeaZf86e", "3DBKc4ioGnMQLlbGQcFDIO"))
alla <- get_album_tracks("7oI0E3DdTbD85rhMg19GSU")
gilberto <- get_artist_audio_features("gilberto gil")
ecm <- get_playlist_audio_features("", "1zN4nK6oHMo2dei2WWPtSL")
```

After you load these features, the easiest way to look at them is to click on the variable in the Environment tab in the top-right pane of RStudio. That will open a full, sortable, spreadsheet-style summary of your data.

But you can also compute summary statistics with the `summary()` function.

```{r}
ecm |>
  summarise(
    mean_speechiness = mean(speechiness),
    mean_acousticness = mean(acousticness),
    mean_liveness = mean(liveness),
    sd_speechiness = sd(speechiness),
    sd_acousticness = sd(acousticness),
    sd_liveness = sd(liveness),
    median_speechiness = median(speechiness),
    median_acousticness = median(acousticness),
    median_liveness = median(liveness),
    mad_speechiness = mad(speechiness),
    mad_acousticness = mad(acousticness),
    mad_liveness = mad(liveness)
  )
```

### Instructions

Open the [Disney Hits](https://open.spotify.com/playlist/37i9dQZF1DX8C9xQcOrE6T?si=brcY2gRMTEeLA73yteEw_A) playlist on Spotify and load it into R, like the demo examples above. Look at the data, starting in spreadsheet form, and compare 'Let It Go' and 'We Don't Talk about Bruno' to the other tracks on the playlist. Is there anything unusual about them? Be prepared to discuss your findings with the group.

If you have time, try identifying outliers on the playlist -- tracks that have extreme and unusual values on one of the Spotify features -- and listen to them. Do you agree with Spotify's assessment? 

Can you also agree on which track on the playlist is the most typical Disney hit? If so, how did you make your decision? If not, why does your group disagree?


## Breakout 3: Digster Decade Comparisions

In this breakout session, you and your group will try to make the best visualisation you can to show the differences between the top pop hits of recent decades.

### Introduction

Let's start by loading playlists from the recent Grammy and Edison awards.

```{r}
grammy <- get_playlist_audio_features("", "6ufHkRP9eCM6w9xbdfwG0q")
edison <- get_playlist_audio_features("", "37i9dQZF1DX8mnKbIkppDf")
```

Next, let's combine data sets with a labelling variable. This is the strategy I recommend for your portfolios: make a separate Spotify playlist for each category in your corpus, and then combine them with a `category` label like the example below.

```{r}
awards <-
  bind_rows(
    grammy |> mutate(category = "Grammys"),
    edison |> mutate(category = "Edisons")
  )
```

The simplest data visualisation of a single, continuous variable is a histogram.

```{r}
grammy |> ggplot(aes(x = energy)) + geom_histogram(binwidth = 0.1)
```

If you want to compare a continuous variable across categories, you can use *faceting* (also known as *small multiples*), a *boxplot* or a *violin plot*.

```{r}
awards |>
  ggplot(aes(x = energy)) +
  geom_histogram(binwidth = 0.1) +
  facet_wrap(~category)
awards |>
  ggplot(aes(x = category, y = energy)) +
  geom_boxplot()
awards |>
  ggplot(aes(x = category, y = energy)) +
  geom_violin()
```

To compare *two* continuous variables, you can create a scatter plot. If you have a large corpus and the points end up on top of each other, `geom_jitter()` adds some random noise without disturbing the overall pattern.

```{r}
grammy |> ggplot(aes(x = valence, y = energy)) + geom_point() + geom_smooth()
grammy |> ggplot(aes(x = valence, y = energy)) + geom_jitter() + geom_smooth()
```

Good visualisations, of course, have many more components. Remember Tufte's six design principles (2006):

  1. Show comparisons, contrasts, differences.
  2. Show causality, mechanism, explanation, systematic structure.
  3. Show multivariate data; that is, show more than 1 or 2 variables.
  4. Completely integrate words, numbers, images, diagrams.
  5. Thoroughly describe the evidence. Provide a detailed title, indicate the authors and sponsors, document the data sources, show complete measurement scales, point out relevant issues.
  6. Analytical presentations ultimately stand or fall on the quality, relevance, and integrity of their content.

```{r}
awards |>                    # Start with awards.
  mutate(
    mode = ifelse(mode == 0, "Minor", "Major")
  ) |>
  ggplot(                     # Set up the plot.
    aes(
      x = valence,
      y = energy,
      size = loudness,
      colour = mode
    )
  ) +
  geom_point() +              # Scatter plot.
  geom_rug(linewidth = 0.1) + # Add 'fringes' to show data distribution.
  geom_text(                  # Add text labels from above.
    aes(
      x = valence,
      y = energy,
      label = label
    ),
    data = 
      tibble(
        label = c("STEAM WALTZ", "Darkness and Scattered Light"),
        category = c("Edisons", "Grammys"),
        valence = c(0.104, 0.0339),
        energy = c(0.992, 0.00166)
      ),
    colour = "black",         # Override colour (not mode here).
    size = 3,                 # Override size (not loudness here).
    hjust = "left",           # Align left side of label with the point.
    vjust = "center",         # Align vertical center of label with the point.
    nudge_x = 0.02            # Nudge the label slightly right.
  ) +
  facet_wrap(~ category) +    # Separate charts per playlist.
  scale_x_continuous(         # Fine-tune the x axis.
    limits = c(0, 1),
    breaks = c(0, 0.50, 1),   # Use grid-lines for quadrants only.
    minor_breaks = NULL       # Remove 'minor' grid-lines.
  ) +
  scale_y_continuous(         # Fine-tune the y axis in the same way.
    limits = c(0, 1),
    breaks = c(0, 0.50, 1),
    minor_breaks = NULL
  ) +
  scale_colour_brewer(        # Use the Color Brewer to choose a palette.
    type = "qual",            # Qualitative set.
    palette = "Paired"        # Name of the palette is 'Paired'.
  ) +
  scale_size_continuous(      # Fine-tune the sizes of each point.
    trans = "exp",            # Use an exp transformation to emphasise loud.
    guide = "none"            # Remove the legend for size.
  ) +
  theme_light() +             # Use a simpler theme.
  labs(                       # Make the titles nice.
    x = "Valence",
    y = "Energy",
    colour = "Mode"
  )
```

### Instructions

Open the user profile for [QmusicNL](https://open.spotify.com/user/qmusicnl?si=7fcc18670839460d) and look at their public playlists. You will see four decade-based playlists:

  - Q-top 500 van de 80s,
  - Q-top 500 van de 90s,
  - Q-top 500 van de 00s, and
  - Q-top 500 van de 10s.

Choose any two decades you want to compare, and make the best visualisation you can that shows the differences between the two decades. Also choose a member of your group who will be prepared to share their screen at the end of the breakout and show off your results. But above all, *help each other* so that every member of your group leaves with some working code.

Start simple, and gradually add details like the ones in the fancy final example.
